The resurgence of the linear optics interferometer --- recent advances \& applications

### 2017-05-12 paper.tex
More revisions
Section 6 (Experimental implementation) has been split into three
sections: state preparation, linear optical circuit, and
photodetection. This is mainly because I would like to discuss some
theory within those sections.

Figures 2 and 3 have been combined into a single figure, and so has
Figures 4 and 5.

\newcommand{\bra}[1]{\langle#1|}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\op}[2]{\hat{\textbf{#1}}_{#2}}
\newcommand{\dagop}[2]{\hat{\textbf{#1}}_{#2}^\dag}
\usepackage{graphicx}

@@

Technical advancements in linear optics have been made on many fronts that make it increasingly attractive for quantum computing applications. It is possible now to put single-photon sources and detectors together with linear-optical networks on a silica chip. The advantage of using such integrated photonics over bulk optics is that it is more stable against phase fluctuations, and miniaturized. This increases the scalability of optical implementations of quantum information protocols. Having all components on chip reduces coupling losses which would be crucial for fault-tolerant quantum computing. With the advantage of potentially piggybacking on existing communication infrastructure, and the relative ease of transporting quantum states of light, as compared to say trapped ions or superconducting qubits, linear optics forms a powerful platform for quantum information processing.

In the last decade, single photons have been produced mainly through spontaneous parametric down-conversion (SPDC), a nonlinear optical process. SPDC produces single photons in correlated pairs that has been the workhorse for providing entanglement in quantum states of light. Through development and improvement in production schemes, the number of high quality indistinguishable photons that can be simultaneously produced have increased steadily over the years. Single-photon pairs, once a novelty, are now run-of-the-mill business for quantum optical labs, and in the last five years, three photons have become the standard benchmark. Experiments with up to ten photons have now been demonstrated. We have also seen the meteoric rise of quantum dots in the role for producing single photons. It is now possible to use a quantum dot to produce multiple single photons which are highly indistinguishable. Quantum dots produce single photons on demand, a significant advantage over SPDC which is an inherently probabilistic process.

The degree of control over single photons has also improved. Using advanced techniques and equipment to manipulate their polarization, time-of-arrival, and orbital angular momentum \cite{bib:Tillmann2015,bib:Bozinovic2013, bib:Nicolas2014, bib:Humphreys2013, bib:Donohue2013}, new ways to engineer entangled quantum states and implement quantum operations have been discovered. These expand the toolbox of baseline resources we have for implementing quantum computations, which could prove crucial in the race to achieve quantum supremacy, that is, the demonstration of a computational task with a quantum  device that is not possible with any known algorithm running on an existing classical supercomputer in a reasonable amount of time.

@@

\section{Experimental implementation}\label{sec:expt}

Significant progress has been made in the experimental realisation of linear optics protocols, and have become standard and widespread. We now discuss some of these advances.

\subsection{State preparation}

@@

\subsubsection{Single-photons}\label{sec:singlephotons}

@@

Some commonly used nonlinear crystals for SPDC are beta barium borate (BBO), periodically poled lithium niobate (PPLN), and periodically poled  potassium titanyl phosphate (PPKTP). Recently, techniques using bismuth triborate (BiBO) have improved to the extent of becoming one of the record-holders in single-photon production \cite{bib:WangChen16}.

Solid-state single photon sources come from semiconducting nanostructures and nitrogen valencies (NV) centers in diamond. Both types are versatile and efficient sources of single photons, however, the photons they produce have suffered from the lack of indistinguishability that is necessary for typical quantum information processing applications. Recent developments in the former have been very promising. Notably, much progress has been made using resonant excitation of quantum dots. Here, laser pulses are used to excite their electronic resonance and trigger the emission of single photons \cite{bib:Muller07,bib:Vamivakas09, bib:Flagg09,bib:Ates09, bib:Dirk10,bib:1748,bib:Jayakumar13,bib:Wei14,bib:Muller14,bib:Unsleber15,
bib:237403,bib:Sweeney14}. An architecture that embeds the quantum dot in a micropillar cavity with the same resonant frequency as the dot exploits the Purcell effect to achieve higher single-photon production efficiency€ \cite{bib:Ding16,bib:213601}. Such resonant excitation of quantum dots overcomes the homogenous broadening of the excited state that causes degradation of photon purity and improving indistinguishability. Using a time-bin loop-based architecture \cite{bib:Motes14} on a stream of single photons produced by such a quantum dot, the same group was able to demonstrate quantum interference between five indistinguishable photons \cite{bib:WangChen16}.

\subsubsection{Einstein-Podolsky-Rosen (EPR) pairs}

@@

\subsubsection{Greenberger-Horne-Zeilinger (GHZ) states}

@@

\subsubsection{Cluster states}\label{section:cluster}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.3\columnwidth]{cluster_state}
\caption{The representation of cluster states as a graph. Vertices represent qubits initialised into the $\ket{+}$ state, while edges represent the application on {\sc cphase} gates, the ordering of which is irrelevant.} \label{fig:cluster}	
\end{figure}

Unfortunately, implementing {\sc cphase} gates using linear optics is complicated and highly non-deterministic. A major improvement upon this is to use fusion gates - rotated polarising beamsplitters, which implement projections onto the Bell basis. These operations fuse smaller cluster states, represented using polarisation encoding, into larger ones, consuming one (type-I fusion) or two (type-II fusion) photons in the process. These operations are shown in Fig.~\ref{fig:fusion}. Although these operations are non-deterministic with a success probability of $1/2$, they require only a single beamsplitter, already a major improvement over directly implementing {\sc cphase} gates. Furthermore, unlike {\sc cphase} gates, fusion operations require only high Hong-Ou-Mandel visibility, rather than the Mach-Zehnder stability required within existing {\sc cphase} gate implementations, a major experimental simplification.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.4\columnwidth]{fusion}
\caption{Fusion gates for joining smaller cluster states into larger ones. Both require only a single polarising beamsplitter, and waveplates. The type-I and -II fusion gates consume 1 or 2 qubits (photons) respectively, creating an edge between the remaining graphs. The type-I gate consumes one fewer photon, but requires number-resolved detection on the detected mode. The type-II gate requires only on/off photodetection, but consumes an additional photon. Thus, the type-I gate can be employed to fuse two Bell pairs into a 3-qubit cluster state, whereas the type-II gate will only grow clusters when beginning with at least 3 qubits per cluster.} \label{fig:fusion}	
\end{figure}

\subsection{Linear optics networks}

@@

\subsubsection{Bulk-optics}

@@

\subsubsection{Waveguides}

@@

\subsubsection{Time-bins}\label{sec:timebins}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.4\columnwidth]{full_architecture}
\caption{A fibre-loop architecture for implementing arbitrary linear optics operations upon a time-bin-encoded pulse-train.
%The source prepares the temporally encoded pulse-train, with time-bin 
%separation $\tau$.
%The pulse-train then enters a dual-loop configuration. The inner loop has 
%%length exactly $\tau$, while the outer one has length $>n\tau$ ($n$ is 
%the number of optical modes). The architecture is controlled via three 
%dynamically controlled beamsplitters. The first and last need only be on/
%off switches, whose sole purpose is to couple in the prepared pulse-
%train, keep it within the outer loop for the required duration, and then 
%couple out of the outer loop, yielding the transformed pulse-train. The 
%central beamsplitter must be able to implement arbitrary classically-
%controlled beamsplitter operations.
} \label{fig:FL_arch}	
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.6\columnwidth]{PBS_TB}
\caption{Mapping between two polarisation-encoded qubits undergoing a polarising beamsplitter (PBS) operation, and its equivalent representation using 4 time-bins in a pulse-train. The PBS completely reflects (transmits) the vertical (horizontal) polarisations. The evolution of the four logical basis states, and their respective outputs, are shown explicitly. Writing out this PBS transformation in matrix form yields a permutation. Taking this permutation and relabelling the modes, we obtain the time-bin transformation shown underneath -- a simple swap of two of the four time-bins.} \label{fig:PBS_TB}	
\end{figure}

@@

\subsection{Photodetection}

@@

\subsection{Linear optical quantum computation (LOQC)}\label{sec:LOQC}
The advancements in linear optics discussed in Section \ref{sec:expt}, quantum computation using linear optics have opened up avenues for implementation. One of the key requirements is nonlinear couplings between the different optical modes. Photons do not naturally interact with one another so some careful engineering is needed to achieve this coupling. Methods used include measurements and feedforwarding \cite{bib:KLM01}, and photon-atom interactions \cite{bib:Brod16}. All models for LOQC also require non-deterministic operations, which succeeds only sometime and have to be repeated. This can be challenging given the limitations of the optical systems. Comprehensive reviews have been written previously to cover this topic \cite{bib:JW12,bib:Kok05}, hence we will focus on recent advances.

 Integrated photonic circuits has greatly improved the feasibility of LOQC. Their small size and interferometric stability enable the requisite quantum interference to happen in certain LOQC settings. Programmable circuits that are reusable have also emerged \cite{bib:Metcalf14,bib:Carolan15}. This saves resources and lead time for producing new circuits. Combining these chips with existing higher-efficiency sources and detectors will expand their capabilities. Loss can be mitigated by moving all components onto the same chip as the linear optical circuit \cite{bib:Sprengers11, bib:Silverstone14}.  What remains is to have fast feedforward also on chip. This might pose the greatest challenge yet for universal LOQC. Only a handful of experiments have been able to implement adaptive feedforward for the purposes of LOQC \cite{bib:Prevedel07,bib:Xiao-Song12,bib:Mikova12,bib:Zhao14}. The bottleneck is in the speed of modulation in the feedforward circuitry.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.5\columnwidth]{MORDOR}
\caption{Architecture of the quantum Fourier transform interferometer for metrology using single-photon states. QFT: Quantum Fourier transform circuit, and $\varphi_k=(k-1)\varphi$. Single photons are injected into the input modes on the left-hand-side, and a $N$-fold coincidence measurement is conducted at the output on the right-hand-side.} \label{fig:MORDOR}	
\end{figure}

>>>>

\newcommand{\bra}[1]{\langle#1|}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\op}[2]{\hat{\textbf{#1}}_{#2}}
\newcommand{\dagop}[2]{\hat{\textbf{#1}}_{#2}^\dag}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

@@

Technical advancements in linear optics have been made on many fronts that make it increasingly attractive for quantum computing applications. It is possible now to put single-photon sources and detectors together with linear-optical networks on a silica chip. The advantage of using such integrated photonics over bulk optics is that it is more stable against phase fluctuations, and miniaturized. This increases the scalability of optical implementations of quantum information protocols. Having all components on chip reduces coupling losses which would be crucial for fault-tolerance. An optical implementation of quantum protocols can potentially piggyback on existing communication infrastructure. Combined withthe relative ease of transporting quantum states of light, as compared to say trapped ions or superconducting qubits, linear optics forms a powerful platform for quantum information processing.

In the last decade, single photons have been produced mainly through a nonlinear optical process know as spontaneous parametric down-conversion (SPDC), a nonlinear optical process. SPDC produces single photons in correlated pairs that has been a staple source of entanglement. Through developments in production, the number of high quality indistinguishable photons that can be simultaneously produced have increased steadily over the years. Single-photon pairs, once a novelty, are now run-of-the-mill business for quantum optical labs, and three photons have become the standard benchmark. Experiments using up to ten photons have been demonstrated. We have also seen the meteoric rise of quantum dots in the role for producing single photons. It is now possible to use a quantum dot to produce multiple single photons which are highly indistinguishable. Quantum dots produce single photons on demand, a significant advantage over SPDC which is an inherently probabilistic process.

Another approach to linear optics is to pack more features known as degrees of freedom onto a single photon, and to increase the degree of control over them. It is possible to manipulate degrees of freedom like polarization, time-of-arrival, and orbital angular momentum \cite{bib:Tillmann2015,bib:Bozinovic2013, bib:Nicolas2014, bib:Humphreys2013, bib:Donohue2013}. Such methods open up new ways to engineer quantum states and quantum operations, and expand the toolbox of baseline resources we have for implementing quantum computations, which could prove crucial in the race to achieve quantum supremacy.

@@

%\section{Experimental implementation}\label{sec:expt}

%Significant progress has been made in the experimental realisation of 
%linear optics protocols, and have become standard and widespread. We now 
%discuss some of these advances.

\section{State preparation}

@@

\subsection{Fock states}\label{sec:singlephotons}

@@

%Some commonly used nonlinear crystals for SPDC are beta barium borate 
%(BBO), periodically poled lithium niobate (PPLN), and periodically poled  
%potassium titanyl phosphate (PPKTP). Recently, techniques using bismuth 
%triborate (BiBO) have improved to the extent of becoming one of the record-
%holders in single-photon production \cite{bib:WangChen16}.

Solid-state single photon sources come from semiconducting nanostructures and nitrogen valencies (NV) centers in diamond. Both types are versatile and efficient sources of single photons, however, the photons they produce have suffered from the lack of indistinguishability that is necessary for typical quantum information processing applications. Recent developments in the former have been very promising. Notably, much progress has been made using resonant excitation of quantum dots. Here, laser pulses are used to excite their electronic resonance and trigger the emission of single photons \cite{bib:Muller07,bib:Vamivakas09, bib:Flagg09,bib:Ates09, bib:Dirk10,bib:1748,bib:Jayakumar13,bib:Wei14,bib:Muller14,bib:Unsleber15,
bib:237403,bib:Sweeney14}. An architecture that embeds the quantum dot in a micropillar cavity uses a resonant effect to achieve higher production efficiency€ and better photon purity \cite{bib:Ding16,bib:213601}. Using a time-bin loop-based architecture \cite{bib:Motes14} on a stream of such photons, quantum interference was demonstrated between five indistinguishable photons \cite{bib:WangHe16}.

\subsection{Einstein-Podolsky-Rosen (EPR) pairs}

@@

\subsection{Greenberger-Horne-Zeilinger (GHZ) states}

@@

\subsection{Cluster states}\label{section:cluster}

@@

\begin{figure}
\centering
\begin{subfigure}{.5\columnwidth}
  \centering
  \includegraphics[width=.6\columnwidth]{cluster_state}
  \caption{}
  \label{fig:cluster}
\end{subfigure}%
\begin{subfigure}{.5\columnwidth}
  \centering
  \includegraphics[width=.8\columnwidth]{fusion}
  \caption{}
  \label{fig:fusion}
\end{subfigure}
\caption{(a) The representation of cluster states as a graph. Vertices represent qubits initialised into the $\ket{+}$ state, while edges represent the application on {\sc cphase} gates, the ordering of which is irrelevant. (b) Fusion gates for joining smaller cluster states into larger ones. Both require only a single polarising beamsplitter, and waveplates. The type-I and -II fusion gates consume 1 or 2 qubits (photons) respectively, creating an edge between the remaining graphs. The type-I gate consumes one fewer photon, but requires number-resolved detection on the detected mode. The type-II gate requires only on/off photodetection, but consumes an additional photon. Thus, the type-I gate can be employed to fuse two Bell pairs into a 3-qubit cluster state, whereas the type-II gate will only grow clusters when beginning with at least 3 qubits per cluster. }
\label{fig:test}
\end{figure}

%\begin{figure}[!htb]
%\centering
%\includegraphics[width=0.3\columnwidth]{cluster_state}
%\caption{The representation of cluster states as a graph. Vertices
% represent qubits initialised into the $\ket{+}$ state, while edges 
%represent the application on {\sc cphase} gates, the ordering of which is 
%irrelevant.} \label{fig:cluster}	
%\end{figure}

Unfortunately, implementing {\sc cphase} gates using linear optics is complicated and highly non-deterministic. A major improvement upon this is to use fusion gates - rotated polarising beamsplitters, which implement projections onto the Bell states. These operations fuse smaller cluster states, represented using polarisation encoding, into larger ones, consuming one (type-I fusion) or two (type-II fusion) photons in the process. These operations are shown in Fig.~\ref{fig:fusion}. Although these operations are non-deterministic with a success probability of $1/2$, they require only a single beamsplitter, already a major improvement over directly implementing {\sc cphase} gates. Furthermore, unlike {\sc cphase} gates, fusion operations require only high Hong-Ou-Mandel visibility, rather than the Mach-Zehnder stability required within existing {\sc cphase} gate implementations, a major experimental simplification.

@@

%\begin{figure}[!htb]
%\centering
%\includegraphics[width=0.4\columnwidth]{fusion}
%\caption{Fusion gates for joining smaller cluster states into larger ones. 
%Both require only a single polarising beamsplitter, and waveplates. The 
%type-I and -II fusion gates consume 1 or 2 qubits (photons) respectively, 
%creating an edge between the remaining graphs. The type-I gate consumes one 
%fewer photon, but requires number-resolved detection on the detected mode. 
%The type-II gate requires only on/off photodetection, but consumes an 
%additional photon. Thus, the type-I gate can be employed to fuse two Bell 
%pairs into a 3-qubit cluster state, whereas the type-II gate will only grow 
%clusters when beginning with at least 3 qubits per cluster.} %
%\label{fig:fusion}	
%\end{figure}

\section{Linear optical networks}\label{sec:network}

@@

\subsection{Bulk-optics}

@@

\subsection{Waveguides}

@@

\subsection{Time-bins}\label{sec:timebins}

@@

\begin{figure}
\centering
\begin{subfigure}{.5\columnwidth}
  \centering
  \includegraphics[width=0.9\columnwidth]{full_architecture}
  \caption{}
  \label{fig:FL_arch}
\end{subfigure}%
\begin{subfigure}{0.5\columnwidth}
  \centering
  \includegraphics[width=\columnwidth]{PBS_TB}
  \caption{}
  \label{fig:fusion}
\end{subfigure}
\caption{(a) A fibre-loop architecture for implementing arbitrary linear optics operations upon a time-bin-encoded pulse-train. (b) Mapping between two polarisation-encoded qubits undergoing a polarising beamsplitter (PBS) operation, and its equivalent representation using 4 time-bins in a pulse-train. The PBS completely reflects (transmits) the vertical (horizontal) polarisations. The evolution of the four logical basis states, and their respective outputs, are shown explicitly. Writing out this PBS transformation in matrix form yields a permutation. Taking this permutation and relabelling the modes, we obtain the time-bin transformation shown underneath -- a simple swap of two of the four time-bins.}
\label{fig:PBS_TB}
\end{figure}

%\begin{figure}[!htb]
%\centering
%\includegraphics[width=0.4\columnwidth]{full_architecture}
%\caption{A fibre-loop architecture for implementing arbitrary linear optics 
%operations upon a time-bin-encoded pulse-train.
%The source prepares the temporally encoded pulse-train, with time-bin 
%separation $\tau$.
%The pulse-train then enters a dual-loop configuration. The inner loop has 
%%length exactly $\tau$, while the outer one has length $>n\tau$ ($n$ is 
%the number of optical modes). The architecture is controlled via three 
%dynamically controlled beamsplitters. The first and last need only be on/
%off switches, whose sole purpose is to couple in the prepared pulse-
%train, keep it within the outer loop for the required duration, and then 
%couple out of the outer loop, yielding the transformed pulse-train. The 
%central beamsplitter must be able to implement arbitrary classically-
%controlled beamsplitter operations.
%} \label{fig:FL_arch}	
%\end{figure}

@@

%\begin{figure}[!htb]
%\centering
%\includegraphics[width=0.6\columnwidth]{PBS_TB}
%\caption{Mapping between two polarisation-encoded qubits undergoing a 
%polarising beamsplitter (PBS) operation, and its equivalent representation 
%using 4 time-bins in a pulse-train. The PBS completely reflects (transmits) 
%the vertical (horizontal) polarisations. The evolution of the four logical 
%basis states, and their respective outputs, are shown explicitly. Writing 
%out this PBS transformation in matrix form yields a permutation. Taking 
%this 
%permutation and relabelling the modes, we obtain the time-bin 
%transformation 
%shown underneath -- a simple swap of two of the four time-bins.} 
%\label{fig:PBS_TB}	
%\end{figure}

@@

\section{Photodetection}

@@

\subsection{Linear optical quantum computation (LOQC)}\label{sec:LOQC}
The advancements in linear optics have opened up avenues for implementation. One of the key requirements is nonlinear couplings between the different optical modes. Photons do not naturally interact with one another so some careful engineering is needed to achieve this coupling. Methods used include measurements and feedforwarding \cite{bib:KLM01}, and photon-atom interactions \cite{bib:Brod16}. The former requires non-deterministic operations, which succeed only part of the time and have to be repeated. This can be challenging given the limited resources to begin with. Comprehensive reviews have been written previously to cover this topic \cite{bib:JW12,bib:Kok05}, hence we will focus on recent advances.

 Integrated photonic circuits has greatly improved the feasibility of LOQC. Their small size and interferometric stability enable the requisite quantum interference to happen in certain LOQC settings. Programmable circuits that are reusable have also emerged \cite{bib:Metcalf14,bib:Carolan15}. This saves resources and lead time for producing new circuits. Combining these chips with existing higher-efficiency sources and detectors will expand their capabilities. Loss can be mitigated by moving all components onto the same chip as the linear optical circuit \cite{bib:Sprengers11, bib:Silverstone14}.  What remains is to have fast feedforward also on chip. This might pose the greatest challenge yet for universal LOQC. The bottleneck is in the speed of modulation in the feedforward circuitry; only a handful of experiments have been able to implement adaptive feedforward for the purposes of LOQC \cite{bib:Prevedel07,bib:Xiao-Song12,bib:Mikova12,bib:Zhao14}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.4\columnwidth]{MORDOR}
\caption{Architecture of the quantum Fourier transform interferometer for metrology using single-photon states. QFT: Quantum Fourier transform circuit, and $\varphi_k=(k-1)\varphi$. Single photons are injected into the input modes on the left-hand-side, and a $N$-fold coincidence measurement is conducted at the output on the right-hand-side.} \label{fig:MORDOR}	
\end{figure}
